EIP CRE transmission network manuscript figures and data

### 2023-01-21 manuscript-figures.Rmd
Update manuscript figures

```{r}
# load libraries
library(tidyverse)
library(ggtree)
library(pheatmap)
library(igraph)
library(regentrans)
library(tidygraph)
library(ggraph)
library(exact2x2)
library(cowplot)
library(readxl)
library(ape)
library(phytools)
library(ggtree)
library(ggtext)
library(lubridate)

@@

```{r}
# load in data
# facility and ST
facil_st <- read_csv('data/isolate_data.csv')
facil_st_mn <- facil_st %>% filter(State == 'MN')
write_csv(facil_st_mn, 'data/mn_isolate_data.csv')
# public isolate data
metadat_complete <- read_csv('data/public_isolate_metadata.csv')
# public isolate trees
tree_list <- read_rds('data/tree_list.rds')
# pairwise snv distances
kp_dists <- read_rds('data/kp_dists.rds')
ec_dists <- read_rds('data/eh_dists.rds')
# phylogenetic trees
tr_kp <- read.tree('data/kp_tr.tree')
tr_ec <- read.tree('data/eh_tr.tree')
# clinical factors
pt_feats_ct <- read_csv('data/pt_feats_ct.csv')
pt_feats_mn <- read_csv('data/pt_feats_mn.csv')
pt_feats_tn <- read_csv('data/pt_feats_tn.csv')
# shared facility exposures
pr_hosp_pairs_ct <- read_csv('data/pr_hosp_pairs_ct.csv')
pr_hosp_pairs_mn <- read_csv('data/pr_hosp_pairs_mn.csv')
pr_hosp_pairs_tn <- read_csv('data/pr_hosp_pairs_tn.csv')
# patient transfer networks
ct_trans_net_sub <- read_csv('data/trans_net_ct.csv')
mn_trans_net_sub <- read_csv('data/trans_net_mn.csv')
tn_trans_net_sub <- read_csv('data/trans_net_tn.csv')
# facility types
facil_types <- read_csv('data/facil_types.csv')
# facility counts
facil_type_counts <- read_csv('data/facil_type_counts.csv')
# facility w prior hospitals
facil_w <- read_csv('data/facil_w_st171_prior_hosps.csv')
# prior tn hospitals
prior_hosps_tn <- read_csv('data/pr_hosps_tn.csv')
# tn facil clutser
tn_facil_cluster <- read_csv('data/tn_facil_cluster.csv')
```

@@

### Species counts by state (Table 1)

@@

facil_st %>% 
  mutate(State = factor(State, levels = c('MN','TN','CT'))) %>% 
  group_by(State, Species) %>% 
  count() %>% full_join(state_dates, .) %>% 
  suppressMessages()
```

### Species counts by facility (Figure S1)

```{r}
fs1 <- facil_st %>% filter(Species != 'Other') %>% 
    mutate(ST = factor(ST, levels = rev(names(st_cols))),
           Species = factor(Species, levels = c('K. pneumoniae','E. hormaechei','Other'))) %>% 
  ggplot(aes(x = Facility_ID, fill = ST)) + geom_bar(color = 'white') + facet_grid(Species~State, scales = 'free', space = 'free') +
  scale_fill_manual(values = st_cols, 
                    breaks = c('E. hormaechei ST171', 'E. hormaechei ST114',
                               'K. pneumoniae ST258', 'K. pneumoniae ST307', 
                               'Other'), 
                    labels = c('*E. hormaechei* ST171', '*E. hormaechei* ST114',
                               '*K. pneumoniae* ST258', '*K. pneumoniae* ST307', 
                               'Other'),
                    drop = FALSE) +
  labs(y = 'Number of isolates', fill = '', x = 'Facility') +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.text.y = element_text(face = 'italic'), legend.text = element_markdown())

ggsave(plot = fs1, filename = 'figures/FigS1.png', width = 7, height = 3)

@@

## Public data (Figure S2)

@@

plot_grid(public_dates_plot, public_loc_plot, ncol = 1, rel_heights = c(0.5, 1), align = 'vh', labels = 'AUTO') %>% 
  ggsave(filename = 'figures/FigS2.png', width = 7, height = 10)
```

## Importation vs. HGT

@@

@@

  grps <- data.frame(id=names(trans_edge_nums),
                     class=trans_sum[trans_edge_nums],
                     cluster=trans_edge_nums) %>% 
    mutate(state = factor(gsub('_.*','',id), levels = c('MN','TN','CT')),
           ST = sapply(id, function(x){
             val <- unlist(unique(metadat_complete$ST[metadat_complete$isolate_no == x & !is.na(metadat_complete$ST)]))
             val <- ifelse(length(val) == 0, paste0('ST', unlist(unique(metadat_complete$mlst[metadat_complete$isolate_no == x & !is.na(metadat_complete$mlst)]))), val)
             val <- ifelse(length(val) == 0 | val == 'ST', st, val)
             ifelse(length(val) == 0, NA, val)
           }),
           species = sapply(id, function(x){
             val <- unlist(unique(metadat_complete$species[metadat_complete$isolate_no == x & !is.na(metadat_complete$species)]))
             ifelse(length(val) == 0, NA, val)
           }),
           ST_orig = ST,
           ST = ifelse(ST %in% c('E. hormaechei ST171','E. hormaechei ST114',
                                 'K. pneumoniae ST258','K. pneumoniae ST307', 
                                 'E. cloacae ST171', 'E. cloacae ST114', 
                                 'ST258', 'ST307', 'ST114', 'ST171'),
                       ST, 'Other'),
           tree = st
           )
  if(nrow(grps) == 0) return(NULL)
  grps
}) %>% 
  bind_rows() %>% suppressWarnings()

transitions <- transitions %>% 
  mutate(class = gsub('NA| ', '', class), 
         class = ifelse(is.na(class), '', class),
         hgt_type = ifelse(grepl('gain_kpc', class), 'Unknown', ''),
         hgt_type = ifelse(class == 'gain_kpc', 'In-state', hgt_type),
         class2 = ifelse(toupper(gsub('to_','',class)) != state & !grepl('gain_kpc', class), '', class),
         class2 = ifelse(grepl('gain_kpc', class2), 'Putative\nHGT', class2),
         class2 = ifelse(class2 == '', 'Unknown', class2),
         class2 = ifelse(grepl('to_', class2), 'Putative\nimportation', class2),
         class2 = factor(class2, levels = c('Putative\nimportation','Putative\nHGT','Unknown')),
         ST = ifelse(ST %in% c('ST258','ST307'), paste0('K. pneumoniae ', ST), ST),
         ST = ifelse(ST %in% c('ST171','ST114'), paste0('E. hormaechei ', ST), ST),
         ST = gsub('cloacae', 'hormaechei', ST),
         ST_orig = gsub('cloacae', 'hormaechei', ST_orig)
         )

keepers <- facil_st %>% 
  filter(!(duplicated(Patient_ID) & duplicated(ST))) %>% 
  pull(Sample_ID)
```

### Trees used (Figure S3)

plot_tr_locs <- function(tr, og, title = NULL){
  tr <- root(tr, og)
  tr <- drop.tip(tr, og)
  states <- metadat_complete %>% select(isolate_no, state) %>% deframe()
  states <- states[tr$tip.label]
  kpc <- metadat_complete %>% select(isolate_no, kpc) %>% deframe()
  kpc <- kpc[tr$tip.label]
  no_info <- is.na(states) & is.na(kpc)
  states <- states[!no_info]
  kpc <- kpc[!no_info]
  kpc <- ifelse(kpc == 1, 'Present', 'Absent')
  tr <- keep.tip(tr, union(names(states), names(kpc)))
  states <- c(states, rep(NA, Nnode(tr)))
  kpc <- c(kpc, rep(NA, Nnode(tr)))
  ggtree(tr) + 
    geom_tippoint(aes(col = states, shape = kpc)) + 
    scale_color_manual(values = state_cols) + 
    scale_shape_manual(values = c('Present' = 16, 'Absent' = 1)) +
    guides(color = guide_legend(ncol = 2)) +
    labs(title = title, shape = 'KPC', col = 'State')
}

tr_st114 <- plot_tr_locs(tree_list[[7]], 'MN_CRE226', '*E. hormaechei* ST114') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st171 <- plot_tr_locs(tree_list[[4]], 'MN_CRE226', '*E. hormaechei* ST171') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st258 <- plot_tr_locs(tree_list[[1]], '573.1451', '*K. pneumoniae* ST258') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st307 <- plot_tr_locs(tree_list[[12]], '573.13662', '*K. pneumoniae* ST307') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
leg <- get_legend(plot_tr_locs(tree_list[[7]], 'MN_CRE226', 'E. hormaechei ST114'))

fs3 <- plot_grid(plot_grid(tr_st171, tr_st114, tr_st258, tr_st307), leg)

ggsave(plot = fs3, filename = 'figures/FigS3.png', width = 8, height = 8)

### Importation vs. HGT (Figure 1)

```{r}
clust_cols <- st_cols[transitions$ST]
names(clust_cols) <- paste0(transitions$state, transitions$ST, transitions$cluster)

clust_cols <- (c(clust_cols[clust_cols == "grey"],
                clust_cols[clust_cols == "coral3"],
                clust_cols[clust_cols == "coral4"],
                clust_cols[clust_cols == "aquamarine3"],
                clust_cols[clust_cols == "aquamarine4"]
                ))

transitions <- transitions %>% filter(id %in% keepers) %>% 
  mutate(ST = factor(ST, levels = (c("E. hormaechei ST171", "E. hormaechei ST114", 
                                     "K. pneumoniae ST258", "K. pneumoniae ST307", 
                                     'Other'))),
         state = factor(state, levels = c('CT', 'MN', 'TN'))) %>%
  arrange(ST)

f1 <- transitions %>% 
  ggplot(aes(x = class2, 
             fill = factor(paste0(state, ST, cluster), 
                           levels = unique(paste0(state, ST, cluster))))) + 
  geom_bar(color = 'white', position = position_stack(reverse = TRUE)) + 
  scale_fill_manual(values = clust_cols,
                    breaks = rev(names(clust_cols)[!duplicated(clust_cols)]),
                    labels = c('*E. hormaechei* ST171', '*E. hormaechei* ST114',
                               '*K. pneumoniae* ST258', '*K. pneumoniae* ST307',
                               'Other'),
                    drop = FALSE) +
  facet_grid(~state, scales = 'free', space = 'free') +
  labs(fill = 'ST cluster', x = 'Cluster type', y = 'Number of isolates') +
  theme(legend.text = ggtext::element_markdown())

@@

### HGT locations (Figure S4)

```{r}
fs4 <- transitions %>% 
  filter(id %in% keepers & hgt_type != '') %>%
  ggplot(aes(x = hgt_type, 
             fill = factor(paste0(state, ST, cluster), 
                           levels = unique(paste0(state, ST, cluster))))) + 
  geom_bar(color = 'white', position = position_stack(reverse = TRUE)) + 
  scale_fill_manual(values = clust_cols,
                    breaks = rev(names(clust_cols)[!duplicated(clust_cols)]),
                    labels = c('*E. hormaechei* ST171', '*E. hormaechei* ST114',
                               '*K. pneumoniae* ST258', '*K. pneumoniae* ST307',
                               'Other'),
                    drop = FALSE) +
  facet_grid(~state, scales = 'free', space = 'free') +
  labs(fill = 'ST cluster', x = 'HGT location', y = 'Number of isolates') +
  theme(legend.text = ggtext::element_markdown())

ggsave(plot = fs4, filename = 'figures/FigS4.png', width = 6, height = 3)

fs4
```
```{r}
# earliest date for ST114 and ST307 
facil_st %>% 
  filter(ST %in% c('E. hormaechei ST114', 'K. pneumoniae ST307') & State == 'TN') %>% 
  group_by(State, ST) %>% 
  mutate(Cult_Date = mdy(Cult_Date)) %>% 
  slice_min(Cult_Date) %>% 
  select(State, ST, Cult_Date)
```

@@

snv_dat <- bind_rows(kp_snv_dists, ec_snv_dists) %>% filter(state1 == state2) %>% 
  mutate(state1 = factor(state1, levels = c('MN','TN','CT')),
         pair_type = ifelse(is.na(pair_type), 'Unknown', pair_type),
         under_thresh = ifelse(pairwise_dist <= 15, '(10,15]','(15,∞)'),
         under_thresh = ifelse(pairwise_dist <= 10, '(5,10]',under_thresh),
         under_thresh = ifelse(pairwise_dist <= 5, '[0,5]',under_thresh),
         under_thresh = factor(under_thresh, levels = c('[0,5]','(5,10]','(10,15]','(15,∞)')),
         ST_orig = st1,
         ST = ifelse(st1 %in% dominant_sts, st1, 'Other'),
         Species = gsub(' ST.*','',st1),
         ST = gsub('.* ','',ST))
```

@@

### Fraction of intra-facility isolates over time (Figure S5)

@@

fs5 <- frac_intras %>% 
    mutate(ST = ifelse(ST %in% c('E. hormaechei ST114','E. hormaechei ST171', 
                                 'K. pneumoniae ST258', 'K. pneumoniae ST307'), ST, 
                       'Other'),
           ST = factor(ST, levels = c('E. hormaechei ST171', 'E. hormaechei ST114', 
                                      'K. pneumoniae ST258', 'K. pneumoniae ST307', 
                                      'Other'),
                       labels = c('*E. hormaechei* ST171', '*E. hormaechei* ST114', 
                                      '*K. pneumoniae* ST258', '*K. pneumoniae* ST307', 
                                      'Other'))) %>% 
  filter(frac_intra != 0 & ST != 'Other') %>% 
  ggplot(aes(x = pairwise_dist, y = frac_intra)) + 
  geom_bar(stat = "identity", fill = "blue", alpha = 0.5) + 
  labs(x = "Pairwise SNV distance", y = "Fraction of intra-facility pairs") + xlim(0, 50) + 
  facet_wrap(state~ST, scales = 'free') +
  theme(strip.text = element_markdown())

ggsave(plot = fs5, filename = 'figures/FigS5.png', width = 6, height = 5)

fs5
```

### Pairwise SNV distances by pair type, state, and dominant STs (Figure 2)

```{r}
f2 <- snv_dat %>% 
  mutate(pairwise_dist = ifelse(pairwise_dist == 0, 1, pairwise_dist),
         pair_type = factor(gsub(' pair', '', pair_type), levels = c('Intra-facility','Inter-facility')),
         Species = gsub('^', '\\*', Species),
         Species = gsub('$', '\\*', Species),
         state1 = factor(state1, levels = c('CT', 'MN', 'TN'))) %>% 
  filter(pair_type != 'Unknown' & ST != 'Other') %>% 
  ggplot(aes(x = pair_type, y = pairwise_dist)) + 
  geom_jitter(aes(fill = under_thresh, color = under_thresh),alpha = 0.5) +
  scale_color_manual(values = c('[0,5]' = "#9EBCDA",'(5,10]' = "#BFD3E6", '(10,15]' = "#E0ECF4",'(15,∞)' = "#FBB4AE")) +
  geom_violin(position = 'identity',alpha=0, color = 'azure4') +
  facet_grid(Species+ST~state1, scales = 'free') + scale_y_log10() +
  labs(x = 'Pair type',y = 'Pairwise SNV distance', color = 'SNV bin', fill = 'SNV bin') +
  theme(strip.text.y = element_markdown())

ggsave(plot = f2, filename = 'figures/Fig2.png', width = 6.5, height = 5)

### Pairwise SNV distances by pair type, state, and all STs (Figure S6)

```{r}
fs6 <- snv_dat %>% filter(state1 == state2 & pair_type != 'Unknown') %>% 
  mutate(state1 = factor(state1, levels = c('CT','MN','TN')),
         pair_type = ifelse(is.na(pair_type), 'Unknown', pair_type),
         st1 = gsub('^', '\\*', st1),
         st1 = gsub('e ', 'e\\* ', st1),
         st1 = gsub('i ', 'i\\* ', st1)) %>% 
  ggplot(aes(x = pairwise_dist, y = st1, color = pair_type)) + 
  geom_vline(xintercept = 10, color = 'darkgrey') +
  geom_jitter(alpha=0.25) + facet_grid(~state1, scales = 'free') +
  scale_color_manual(values = c('lightslateblue','black')) +
  labs(y = '', x = 'Pairwise SNV distance', color = '') +
  theme(axis.text.y = element_markdown())

ggsave(plot = fs6, filename = 'figures/FigS6.png', width = 7, height = 5)

fs6
```

### Patristic distance sensitivity analysis (Figure S7)

>>>>

```{r}
# load libraries
library(tidyverse)
library(ggtree)
library(pheatmap)
library(igraph)
library(regentrans)
library(tidygraph)
library(ggraph)
library(exact2x2)
library(cowplot)
library(readxl)
library(ape)
library(phytools)
library(ggtree)
library(ggtext)
library(lubridate)
library(scales)
library(ggrepel)

@@

```{r}
# load in data
# facility and ST
facil_st_orig <- read_csv('data/isolate_data.csv') %>% 
  mutate(cult_year = gsub('-.*|.*/', '', Cult_Date))
# to make analyses more consistent across states
facil_st <- facil_st_orig %>% 
  filter(cult_year >= 2016)
# public isolate data
metadat_complete <- read_csv('data/public_isolate_metadata.csv') 
# public isolate trees
tree_list <- read_rds('data/tree_list.rds')
# pairwise snv distances
kp_dists <- read_rds('data/kp_dists.rds')
ec_dists <- read_rds('data/eh_dists.rds')
# phylogenetic trees
tr_kp <- read.tree('data/kp_tr.tree')
tr_ec <- read.tree('data/eh_tr.tree')
# clinical factors
pt_feats_ct <- read_csv('data/pt_feats_ct.csv')
pt_feats_mn <- read_csv('data/pt_feats_mn.csv')
pt_feats_tn <- read_csv('data/pt_feats_tn.csv')
# shared facility exposures
pr_hosp_pairs_ct <- read_csv('data/pr_hosp_pairs_ct.csv')
pr_hosp_pairs_mn_orig <- read_csv('data/pr_hosp_pairs_mn.csv')
pr_hosp_pairs_mn <- pr_hosp_pairs_mn_orig
pr_hosp_pairs_tn <- read_csv('data/pr_hosp_pairs_tn.csv')
# patient transfer networks
ct_trans_net_sub <- read_csv('data/trans_net_ct.csv')
mn_trans_net_sub <- read_csv('data/trans_net_mn.csv')
tn_trans_net_sub <- read_csv('data/trans_net_tn.csv')
# facility types
facil_types <- read_csv('data/facil_types.csv')
# facility counts
facil_type_counts <- read_csv('data/facil_type_counts.csv')
# prior tn hospitals
prior_hosps_tn <- read_csv('data/pr_hosps_tn.csv')
# tn facil clutser
tn_facil_cluster <- read_csv('data/tn_facil_cluster.csv')
```

@@

### Isolate counts by state

@@

facil_st %>% 
  mutate(State = factor(State, levels = c('MN','TN','CT'))) %>% 
  group_by(State) %>% 
  count() %>% full_join(state_dates, .) %>% 
  suppressMessages()
```

### Isolate overview (Figure S1A)

```{r}
fs1a <- facil_st %>% 
  mutate(ST = ifelse(ST %in% c('E. hormaechei ST114','E. hormaechei ST171', 
                                 'K. pneumoniae ST258', 'K. pneumoniae ST307'), ST, 
                       'Other'),
           ST = factor(ST, levels = rev(c('E. hormaechei ST171', 'E. hormaechei ST114', 
                                      'K. pneumoniae ST258', 'K. pneumoniae ST307', 
                                      'Other')))) %>% 
  group_by(State, ST, cult_year) %>% 
  filter(!is.na(cult_year)) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(x = cult_year, y = count, fill = ST)) +
  facet_grid(~State, scales = 'free', space = 'free') +
  geom_col() +
  scale_fill_manual(values = st_cols, 
                    breaks = c('E. hormaechei ST171', 'E. hormaechei ST114',
                               'K. pneumoniae ST258', 'K. pneumoniae ST307',
                               'Other'), 
                    labels = c('KPC-Eh ST171', 'KPC-Eh ST114',
                               'KPC-Kp ST258', 'KPC-Kp ST307',
                               'Other')) +
  theme(legend.text = element_markdown()) +
  labs(x = 'Culture year', y = 'Number of isolates')
```

### Species counts by facility (Figure S1B)

```{r}
fs1b <- facil_st %>% filter(Species != 'Other') %>% 
    mutate(ST = factor(ST, levels = rev(names(st_cols))),
           Species = factor(Species, levels = c('K. pneumoniae','E. hormaechei','Other'),
                            labels = c('KPC-Kp', 'KPC-Eh', 'Other'))) %>% 
  ggplot(aes(x = Facility_ID, fill = ST)) + geom_bar(color = 'white') + 
  facet_grid(Species~State, scales = 'free', space = 'free') +
  scale_fill_manual(values = st_cols, 
                    breaks = c('E. hormaechei ST171', 'E. hormaechei ST114',
                               'K. pneumoniae ST258', 'K. pneumoniae ST307', 
                               'Other'), 
                    labels = c('KPC-Eh ST171', 'KPC-Eh ST114',
                               'KPC-Kp ST258', 'KPC-Kp ST307', 
                               'Other'),
                    drop = FALSE) +
  labs(y = 'Number of isolates', fill = '', x = 'Facility') +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.text = element_markdown())
```

```{r, fig.width=5, fig.height=5}
fs1 <- plot_grid(fs1a, fs1b + theme(legend.position = 'none'), ncol = 1, labels = 'AUTO')

ggsave(plot = fs1, filename = 'figures/FigS1.png', width = 5, height = 5)

@@

## Public data & trees (Figure S2)

@@

fs2ab <- plot_grid(public_dates_plot, public_loc_plot, ncol = 1, rel_heights = c(0.5, 1), align = 'vh', axis = 'trbl', labels = 'AUTO') 
```

plot_tr_locs <- function(tr, og, title = NULL){
  tr <- root(tr, og)
  tr <- drop.tip(tr, og)
  states <- metadat_complete %>% select(isolate_no, state) %>% deframe()
  states <- states[tr$tip.label]
  kpc <- metadat_complete %>% select(isolate_no, kpc) %>% deframe()
  kpc <- kpc[tr$tip.label]
  no_info <- is.na(states) & is.na(kpc)
  states <- states[!no_info]
  kpc <- kpc[!no_info]
  kpc <- ifelse(kpc == 1, 'Present', 'Absent')
  tr <- keep.tip(tr, union(names(states), names(kpc)))
  states <- c(states, rep(NA, Nnode(tr)))
  states <- factor(states, levels = sort(names(state_cols[!is.na(names(state_cols))])))
  kpc <- c(kpc, rep(NA, Nnode(tr)))
  ggtree(tr) + 
    geom_tippoint(aes(col = states, shape = kpc)) + 
    scale_color_manual(values = state_cols, drop=FALSE) + 
    scale_shape_manual(values = c('Present' = 16, 'Absent' = 1)) +
    guides(color = guide_legend(ncol = 3)) +
    labs(title = title, shape = 'KPC', col = 'State')
}

tr_st114 <- plot_tr_locs(tree_list[[7]], 'MN_CRE226', '*E. hormaechei* ST114') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st171 <- plot_tr_locs(tree_list[[4]], 'MN_CRE226', '*E. hormaechei* ST171') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st258 <- plot_tr_locs(tree_list[[1]], '573.1451', '*K. pneumoniae* ST258') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
tr_st307 <- plot_tr_locs(tree_list[[12]], '573.13662', '*K. pneumoniae* ST307') + 
  theme(legend.position = 'none', plot.title = ggtext::element_markdown())
leg <- get_legend(plot_tr_locs(tree_list[[1]], '573.1451', '*K. pneumoniae* ST258'))

fs2c <- plot_grid(plot_grid(tr_st171, tr_st114, tr_st258, tr_st307), 
                  leg, ncol = 1, rel_heights = c(1, 0.6))
```

```{r, fig.width=10, fig.height=8}
fs2 <- plot_grid(fs2ab, fs2c, labels = c('', 'C'), label_x = -0.1)
ggsave('figures/FigS2.png', width = 10, height = 8)
fs2
```

## Evidence of importation

@@

keepers <- facil_st %>% 
  filter(!(duplicated(Patient_ID) & duplicated(ST))) %>% 
  pull(Sample_ID)

@@

  grps <- data.frame(id=names(trans_edge_nums),
                     class=trans_sum[trans_edge_nums],
                     cluster=trans_edge_nums) %>% 
    mutate(state = factor(gsub('_.*','',id), levels = c('MN','TN','CT')),
           ST = sapply(id, function(x){
             val <- unlist(unique(metadat_complete$ST[metadat_complete$isolate_no == x & !is.na(metadat_complete$ST)]))
             val <- ifelse(length(val) == 0, paste0('ST', unlist(unique(metadat_complete$mlst[metadat_complete$isolate_no == x & !is.na(metadat_complete$mlst)]))), val)
             val <- ifelse(length(val) == 0 | val == 'ST', st, val)
             ifelse(length(val) == 0, NA, val)
           }),
           species = sapply(id, function(x){
             val <- unlist(unique(metadat_complete$species[metadat_complete$isolate_no == x & !is.na(metadat_complete$species)]))
             ifelse(length(val) == 0, NA, val)
           }),
           ST_orig = ST,
           ST = ifelse(ST %in% c('E. hormaechei ST171','E. hormaechei ST114',
                                 'K. pneumoniae ST258','K. pneumoniae ST307', 
                                 'E. cloacae ST171', 'E. cloacae ST114', 
                                 'ST258', 'ST307', 'ST114', 'ST171'),
                       ST, 'Other'),
           tree = st
           )
  if(nrow(grps) == 0) return(NULL)
  grps
}) %>% 
  bind_rows() %>% suppressWarnings() %>% 
  mutate(class = gsub('NA| ', '', class), 
         class = ifelse(is.na(class), '', class),
         class2 = ifelse(grepl('gain_kpc', class), 'No', class),
         class2 = ifelse(class2 == '', 'No', class2),
         class2 = ifelse(grepl('to_', class2), 'Yes', class2),
         class2 = factor(class2, levels = c('Yes','No')),
         ST = ifelse(ST %in% c('ST258','ST307'), paste0('K. pneumoniae ', ST), ST),
         ST = ifelse(ST %in% c('ST171','ST114'), paste0('E. hormaechei ', ST), ST),
         ST = gsub('cloacae', 'hormaechei', ST),
         ST_orig = gsub('cloacae', 'hormaechei', ST_orig),
         ST = factor(ST, levels = (c("E. hormaechei ST171", "E. hormaechei ST114", 
                                     "K. pneumoniae ST258", "K. pneumoniae ST307", 
                                     'Other'))),
         state = factor(state, levels = c('CT', 'MN', 'TN'))) %>% 
  filter(id %in% keepers) %>% 
  arrange(ST)
```

### Evidence of importation (Figure 1)

```{r}
transitions <- transitions %>% 
  mutate(ST = ifelse(grepl('ST258', ST_orig), 'K. pneumoniae ST258', as.character(ST))) 
clust_cols <- st_cols[as.character(transitions$ST)]
names(clust_cols) <- paste0(transitions$state, transitions$ST, transitions$cluster)

clust_cols <- (c(clust_cols[clust_cols == "grey"][sort(names(clust_cols[clust_cols == "grey"]), decreasing = TRUE)],
                clust_cols[clust_cols == "coral3"][sort(names(clust_cols[clust_cols == "coral3"]), decreasing = TRUE)],
                clust_cols[clust_cols == "coral4"][sort(names(clust_cols[clust_cols == "coral4"]), decreasing = TRUE)],
                clust_cols[clust_cols == "aquamarine3"][sort(names(clust_cols[clust_cols == "aquamarine3"]), decreasing = TRUE)],
                clust_cols[clust_cols == "aquamarine4"][sort(names(clust_cols[clust_cols == "aquamarine4"]), decreasing = TRUE)]
                ))
```

```{r, fig.width=7, fig.height=4}
f1 <- transitions %>% 
  ggplot(aes(x = class2, 
             fill = factor(paste0(state, ST, cluster), 
                           levels = unique(paste0(state, ST, cluster))))) + 
  geom_bar(color = 'white', position = position_stack(reverse = TRUE)) + 
  scale_fill_manual(values = clust_cols,
                    breaks = rev(names(clust_cols)[!duplicated(clust_cols)]),
                    labels = c('KPC-Eh ST171', 'KPC-Eh ST114',
                               'KPC-Kp ST258', 'KPC-Kp ST307',
                               'Other'),
                    drop = FALSE) +
  facet_grid(~state, scales = 'free', space = 'free') +
  labs(fill = 'ST cluster', x = 'Evidence that KPC-harboring strain was imported', y = 'Number of isolates') +
  theme(legend.text = ggtext::element_markdown())

@@

```{r}
# earliest date for ST114 and ST307 
facil_st %>% 
  filter(ST %in% c('E. hormaechei ST114', 'K. pneumoniae ST307') & State == 'TN') %>% 
  group_by(State, ST) %>% 
  mutate(Cult_Date = mdy(Cult_Date)) %>% 
  slice_min(Cult_Date) %>% 
  select(State, ST, Cult_Date)
```

@@

snv_dat <- bind_rows(kp_snv_dists, ec_snv_dists) %>% filter(state1 == state2) %>% 
  mutate(state1 = factor(state1, levels = c('MN','TN','CT')),
         pair_type = ifelse(is.na(pair_type), 'Unknown', pair_type),
         under_thresh = ifelse(pairwise_dist <= 15, '(10,15]','(15,∞)'),
         under_thresh = ifelse(pairwise_dist <= 10, '(5,10]',under_thresh),
         under_thresh = ifelse(pairwise_dist <= 5, '[0,5]',under_thresh),
         under_thresh = factor(under_thresh, levels = c('[0,5]','(5,10]','(10,15]','(15,∞)')),
         ST_orig = st1,
         ST = ifelse(st1 %in% dominant_sts, st1, 'Other'),
         Species = gsub(' ST.*','',st1),
         ST = gsub('.* ','',ST)) %>% 
  left_join(facil_st %>% select(Sample_ID, cult_year) %>% 
              rename(isolate1 = Sample_ID, year1 = cult_year)) %>% 
  left_join(facil_st %>% select(Sample_ID, cult_year) %>% 
              rename(isolate2 = Sample_ID, year2 = cult_year))
```

@@

### Fraction of intra-facility isolates over time (Figure S3A)

@@

fs3a <- frac_intras %>% 
    mutate(ST = ifelse(ST %in% c('E. hormaechei ST114','E. hormaechei ST171', 
                                 'K. pneumoniae ST258', 'K. pneumoniae ST307'), ST, 
                       'Other'),
           ST = factor(ST, levels = c('E. hormaechei ST171', 'E. hormaechei ST114', 
                                      'K. pneumoniae ST258', 'K. pneumoniae ST307', 
                                      'Other'),
                       labels = c('KPC-Eh<br>ST171', 'KPC-Eh<br>ST114', 
                                      'KPC-Kp<br>ST258', 'KPC-Kp<br>ST307', 
                                      'Other'))) %>% 
  filter(frac_intra != 0 & ST != 'Other') %>% 
  filter(pairwise_dist <= 50) %>%
  ggplot(aes(x = pairwise_dist, y = frac_intra, fill = state)) + 
  geom_bar(stat = "identity") + # fill = "blue", 
  geom_vline(xintercept = c(5, 10, 15), alpha = 0.2) +
  labs(x = "Pairwise SNV distance", y = "Fraction of intra-facility pairs", fill = 'State') + 
  facet_grid(ST~., scales = 'free') +
  scale_fill_brewer(palette = 'Set1') +
  scale_x_continuous(breaks = seq(0, 50, by = 5)) +
  theme(strip.text.y = element_markdown(angle = 0), panel.grid = element_blank())
```

### Pairwise SNV distances by pair type, state, and other STs (Figure S3B)

```{r}
fs3b <- snv_dat %>% 
  filter(state1 == state2 & pair_type != 'Unknown' & !is.na(pair_type)) %>%
  mutate(pairwise_dist = ifelse(pairwise_dist == 0, 0.1, pairwise_dist),
         pair_type = factor(gsub(' pair', '', pair_type), levels = c('Intra-facility','Inter-facility')),
         state1 = factor(state1, levels = c('CT','MN','TN')),
         # pair_type = ifelse(is.na(pair_type), 'Unknown', pair_type),
         # pair_type = gsub(' pair', '', pair_type),
         st1 = gsub('K. pneumoniae', 'KPC-Kp', st1),
         st1 = gsub('E. hormaechei', 'KPC-Eh', st1)) %>% 
  ggplot(aes(x = pairwise_dist, y = st1, color = under_thresh)) + 
  # geom_vline(xintercept = c(5, 10, 15), color = 'darkgrey', linetype = 'dashed') +
  geom_jitter(alpha=0.5) + 
  facet_grid(pair_type~state1, scales = 'free', space = 'free') +
  scale_color_manual(values = c('[0,5]' = "#9EBCDA",'(5,10]' = "#BFD3E6", '(10,15]' = "#E0ECF4",'(15,∞)' = "#FBB4AE")) +
  scale_x_continuous(breaks = seq(0, 600, by = 50)) +
  # scale_color_manual(values = c('lightslateblue','black')) +
  labs(y = '', x = 'Pairwise SNV distance', color = 'SNV bin') 
```

### Patristic distance sensitivity analysis (Figure S3C)

```{r}
fs3c <- left_join(snv_dat %>% rename(pd_snv = pairwise_dist), pat_dat %>% rename(pd_pat = pairwise_dist)) %>% 
  filter(state1 == state2 & pair_type != 'Unknown' & !is.na(pair_type)) %>%
  mutate(pair_type = factor(gsub(' pair', '', pair_type), levels = c('Intra-facility','Inter-facility')),
         state1 = factor(state1, levels = c('CT','MN','TN')),
         st1 = gsub('K. pneumoniae', 'KPC-Kp', st1),
         st1 = gsub('E. hormaechei', 'KPC-Eh', st1)) %>% 
  ggplot(aes(x = pd_pat, y = st1, color = under_thresh)) + 
  geom_jitter(alpha=0.5) + 
  facet_grid(pair_type~state1, scales = 'free', space = 'free') +
  scale_color_manual(values = c('[0,5]' = "#9EBCDA",'(5,10]' = "#BFD3E6", '(10,15]' = "#E0ECF4",'(15,∞)' = "#FBB4AE")) +
  scale_x_continuous(breaks = seq(0, 0.01, by = 0.00003)) +
  labs(y = '', x = 'Pairwise patristic distance', color = 'SNV bin') 
```

```{r, fig.width=7.5, fig.height=10.5}
fs3 <- plot_grid(plotlist = list(fs3a, fs3b, fs3c), ncol = 1, labels = 'AUTO',
                 rel_heights = c(0.75, 1.1, 1.1),
                 align = 'hv', axis = 'trbl')

ggsave(plot = fs3, filename = 'figures/FigS3.png', width = 7.5, height = 10.5)

## Intra-facility analysis (Figure 2A)

```{r}
min_dat <- lapply(unique(c(as.character(snv_dat$isolate1), as.character(snv_dat$isolate2))), function(i){
  snv_dat %>% filter((isolate1 == i | isolate2 == i) & pair_type != 'Unknown') %>%
    group_by(pair_type, ST_orig, state1) %>% 
    summarize(dist_min = min(pairwise_dist)) %>% 
    pivot_wider(names_from = pair_type, values_from = dist_min)
}) %>% bind_rows() %>% suppressMessages() %>% suppressWarnings()
```
